---
title: "Hotel Bookings"
author: "Braden Crimmins, Jaylen Burch, Tanner Burke, Derek Fuller, Daniel Hult"
date: "2/15/2021"
output: 
  html_document:
    theme: united
    toc: TRUE
    toc_float:
      collapsed: FALSE
      smooth_scroll: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Cleaning

### General Setup
```{r}
library(tidyverse)
hotels <- read.csv("hotel_bookings.csv")
```

### Redefining Variable Types
The variables listed below are redefined from integers/characters to factors. The purpose of this is to set each variable up to perform certain analyses.
```{r}
# Arrival Date Year
hotels$arrival_date_year <- as.factor(hotels$arrival_date_year) 
# Arrival Date Month
hotels$arrival_date_month <- as.factor(hotels$arrival_date_month)
# Arrival Date Week Number
hotels$arrival_date_week_number <- as.factor(hotels$arrival_date_week_number)
# Arrival Date Day of Month
hotels$arrival_date_day_of_month <- as.factor(hotels$arrival_date_day_of_month)
# Booking Changes
hotels$booking_changes <- as.factor(hotels$booking_changes)
# Required Car Parking Spaces
hotels$required_car_parking_spaces <- as.factor(hotels$required_car_parking_spaces)
# Total of Special Requests
hotels$total_of_special_requests <- as.factor(hotels$total_of_special_requests)
# Country
hotels$country <- as.factor(hotels$country)
# Meal
hotels$meal <- as.factor(hotels$meal)
# Market Segment
hotels$market_segment <- as.factor(hotels$market_segment)
# Distribution Channel
hotels$distribution_channel <- as.factor(hotels$distribution_channel)
# Customer Type
hotels$customer_type <- as.factor(hotels$customer_type)
# Reservation Status
hotels$reservation_status <- as.factor(hotels$reservation_status)
# Reserved Room Type
hotels$reserved_room_type <- as.factor(hotels$reserved_room_type)
# Assigned Room Type
hotels$assigned_room_type <- as.factor(hotels$assigned_room_type)
# Hotel
hotels$hotel <- as.factor(hotels$hotel)
```

### Creating New Columns
The purpose of creating these new columns is to allow us to perform the necessary analyses.
```{r}
# Combines bookings with children and bookings with kids as one entity
hotels$kids <- hotels$children + hotels$babies

# Calculates Total Number of Bookings
hotels$totalbookings <- hotels$previous_cancellations + hotels$previous_bookings_not_canceled

# Calculates Proportion of Cancellations to Total Bookings
hotels$propcanceled <- ifelse(hotels$totalbookings != 0, hotels$previous_cancellations/hotels$totalbookings, NA)

# Creates a boolean value to determine if a booking was made by a customer who has booked in the past
hotels$bookedbefore <- as.factor(ifelse(hotels$totalbookings != 0, 1, 0))

# Determines for all customers if they have historically cancelled more than half of their bookings
hotels$cancelled_morethanhalf <- as.factor(ifelse(hotels$propcanceled > 0.5, 1, 0))
```

### Creating New Variables
Determining the total number of cancellations allows us to plot certain variables against total number of cancellations as a means to visually find any correlations in variables to cancellations.
```{r}
# Total number of cancellations
num_cancel <- sum(hotels$is_canceled)
```

# Customer Analysis

### Bookings by Country
Below is a table and graph showing the number of cancellations made per country.
```{r}
hotels %>% group_by(country) %>% summarise(num_cancel = sum(is_canceled)) %>% arrange(desc(num_cancel)) %>% head(10) %>% print %>% ggplot(aes(x = country, y = num_cancel, fill= num_cancel))+ geom_col()

```
From the data, Portugal is shown to be the country with the most cancellations by a significant amount. However, this is expected given that the hotels from the dataset are located in Portugal, and thus, attract more customers from Portugal than other countries. This chart can double as an indicator of where the customer base for the hotels are located.  

### Cancellations Per Booking Method
Below is a bar chart showcasing the number of cancellations made after booking through different distribution channels.
```{r}
hotels %>% group_by(distribution_channel) %>% arrange(desc(num_cancel)) %>% head(10) %>% ggplot(aes(x = distribution_channel, y = num_cancel, fill = num_cancel)) + geom_col()
```
The output shows us the number of canceled bookings made after booking through certain distribution channels. Customers who booked direct canceled more often than any other distribution channel. Customers who booked through the corporate distribution channel made by far the least number of cancellations. Corporate distribution channels implies business travel and thus, there is an obligation for the customers (company employees) to attend given that the purposes of the booking is for work.

### Cancellations Per Customer Type
Below is a bar chart showcasing the number of cancellations made for different customer types. Customer Types are categorized by Contract, Group, Transient, and Transient-Party. 

###### Contract customers are defined as customers in which the booking has an allotment or other type of contract associated to it.

###### Group customers are defined as customers booking with an associated group.

###### Transient Customers are defined as customers booking as not part of a group or contract, and is not associated with any other transient booking.

###### Transient-Party customers are defined as customers whose booking is transient, but is also associated to at least one other transient booking.
```{r}
hotels %>% group_by(customer_type) %>%
  arrange(desc(num_cancel)) %>%
  ggplot(aes(x = customer_type, y = num_cancel, fill = num_cancel)) + geom_col()
```
From the data, we notice that the number of transient customers who canceled their stay is much higher than any other category. This makes sense intuitively as transient customers have less of a liability when canceling their stay compared to transient-party customers or corporate customers. Transient customers are primarily individuals traveling, and thus, do not have any attached obligations to their booking whereas contract, group, and transient-part customer types have some from of attached obligation to their booking (e.g. other people, company, etc.).

### Reservation Status
Below is a bar chart that shows a comparison between the number of bookings canceled, number of bookings successfully stayed and checked-out, and the number of bookings that resulted in a no-show across the total number of bookings in the data set.
```{r}
hotels %>% group_by(reservation_status) %>%
  ggplot(aes(x = reservation_status, y = num_cancel)) + geom_col()
```
It can be noted that most of the bookings resulted in a successful check-in and check-out. A substantial number of bookings also resulted in a cancellation. However, this presents a unique opportunity for the company to identify and charge customers higher cancellation fees who are likely to result in a cancellation after booking. 

### Cancellations Per Booking Month
Below is a table and bar chart showcasing the number of cancellations made per month from July 2015 to August 2017. 
```{r}
hotels %>% group_by(arrival_date_month) %>% summarise(num_cancel = sum(is_canceled)) %>% arrange(desc(num_cancel)) %>% head(12) %>% print %>% ggplot(aes(x = arrival_date_month, y = num_cancel, fill= num_cancel))+ geom_col()
```
Ultimately, there is high variability in the number of cancellations per month, there does appear to be a trend. The five months with the lowest number of cancellations happen consecutively (November, December, January, February, March). While this trend can be attributed to a possible lower number of bookings made in these five months, there can also be insight into possible seasonality effects occurring in which the months of November, December, January, February, and March yield higher rates of cancellation. 

On the other hand, cancellations appeared to be highest during the summer months such as May, June, July, and August. This could be possibly be attributed to people traveling more during summer months; however, there is also potential insight for seasonality effects mentioned above.

### Children and Cancellations
The table below shows the number of customers who canceled their stay and whether or not they had children booked with them. 
```{r}
hotels %>% select(is_canceled, children) %>% 
  na.omit(children) %>%
  mutate(has_child = ifelse(children != 0, 1, 0)) %>%
  group_by(has_child) %>%
  summarise(num_cancel = sum(is_canceled))
```
Of customers who canceled, 41,094 customers booked/canceled without children and 3,126 customers booked/canceled with children. The total number of bookings made with children was 110,796 whereas the total number of bookings made without children was 8,590. Calculating the cancellation rate between these two segments indicates a 36% cancellation rate for customers booking with children and a 37% cancellation rate for customers booking without children. 

Ultimately, there is no statistical significance in the differences of cancellation rates between these two customer segments; however, this information is insightful given that this is a common metric hotel companies use to evaluate the cancellation rate of customers. 

### Previous Bookings
```{r}
table(hotels$bookedbefore)
round(prop.table(table(hotels$bookedbefore)), 3)

```
These tables show the amount and proportion of customers who have previously booked reservations in the past and whether or not they checked-in or canceled.

```{r}
round(prop.table(table(hotels$is_canceled[hotels$bookedbefore == 0])), 2)
round(prop.table(table(hotels$is_canceled[hotels$bookedbefore == 1])), 2)
#if a customer has booked before, they are much more likely to cancel. 1/3 of first-time customers are likely to cancel whereas 2/3 of repeat customers are likely to cancel.

round(prop.table(table(hotels$is_canceled[hotels$cancelled_morethanhalf == 0])), 2)
round(prop.table(table(hotels$is_canceled[hotels$cancelled_morethanhalf == 1])), 2)
# IF a customer has NOT canceled more than half of their bookings, then they will almost certainly check-in/come. IF a customer HAS canceled more than half of their bookings, then they will almost certainly cancel again.

# -----------------------------------------------------------

boxplot(propcanceled ~ is_canceled, data = hotels, na.rm = TRUE)

# table(hotels$is_canceled[is.na(hotels$propcanceled)], hotels$cancelled_morethanhalf, na.rm = TRUE)

table(hotels$is_canceled[!is.na(hotels$propcanceled)], hotels$cancelled_morethanhalf[!is.na(hotels$propcanceled)])

round(prop.table(table(hotels$is_canceled[!is.na(hotels$propcanceled)], hotels$cancelled_morethanhalf[!is.na(hotels$propcanceled)])), 3)

# Only 1/3 of your reservations by people who have made reservations before are made my people who have a history of showing up more than half the time. These people consistently show up and check in to their stay. The remaining 2/3 of reservations are made by people who have a history of canceling. These people consistently book a reservation and then, ultimately, cancel that reservation.

# if people canceled more than half of prev. reservations, it 
```


# Conclusion
```{r}
# Standard factors that a manager may believe to drive cancellation (e.g. hotel type, children/no children) actually show little to no impact on cancellation rates. Rather, the largest indicator is the customers' history in booking/canceling behavior. Customers who have a history of canceling reservations in the past are more likely to do so in the future, therefore they should be charged a higher rate or cancellation fee. Repeat customers who have fulfilled their reservation in the past are most likely to fulfill future reservations, so these individuals should be our focus.


```