---
title: "Project 2"
author: "Braden Crimmins, Jaylen Burch, Tanner Burke, Derek Fuller"
date: "3/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this project, we hope to identify discrepancies between the factors used to decide interest rates on loans and the factors that actually make such loans risky. We will accomplish this by first generating a linear regression model that determines which factors are used in determining interest rates (see "Linear Regression"), and then comparing it to our logistic model that determines which factors actually influence the risk of default (see "Logistic Regression"). By comparing these models, we can identify which factors lenders might incorrectly believe are 'risky', and we can generate profit / make the market more efficient by lending to those underserved populations.

First, we read in and clean our data. Making variables factors ensures our models don't accidentally misinterpret them, and ensures we generate meaningful results. 
```{r}
loans <- read.csv("ProjectA_Listings2013.csv")

loans$loan_status <- as.factor(loans$loan_status)
loans$loan_status_description <- as.factor(loans$loan_status_description)
loans$loan_origination_date <- as.factor(loans$loan_origination_date)
loans$prosper_rating <- as.factor(loans$prosper_rating)
loans$scorex <- as.factor(loans$scorex)
loans$prosper_score <- as.factor(loans$prosper_score)
loans$listing_category_id <- as.factor(loans$listing_category_id)
loans$income_range <- as.factor(loans$income_range)
loans$income_range_description <- as.factor(loans$income_range_description)
loans$income_verifiable <- as.factor(loans$income_verifiable)
loans$employment_status_description <- as.factor(loans$employment_status_description)
loans$occupation <- as.factor(loans$occupation)
loans$borrower_state <- as.factor(loans$borrower_state)
loans$borrower_city <- as.factor(loans$borrower_city)
loans$lender_indicator <- as.factor(loans$lender_indicator)
loans$first_recorded_credit_line <- as.factor(loans$first_recorded_credit_line)
loans$is_homeowner <- as.factor(loans$is_homeowner)
```

## Linear Regression
Once we've read our data, we can begin working on our linear regression. We begin by further cleaning our data so it is as suitable as possible for the linear regression we plan to do.
```{r}
# Create copy of full dataset for the linear model
loans_lm <- loans

# Remove data we wouldn't have when setting rate
loans_lm$loan_status_description <- NULL
loans_lm$number_of_days <- NULL
loans_lm$principal_balance <- NULL
loans_lm$amount_delinquent <- NULL
loans_lm$installment_balance <- NULL
loans_lm$real_estate_balance <- NULL
loans_lm$revolving_balance <- NULL

# Remove duplicative data
loans_lm$income_range_description <- NULL
```

Once this data is cleaned, we can create a linear model with most of the remaining terms (excluding terms like city and state that don't add much information and just make our model infeasible to compute)
```{r}
# Create linear model 
m1 <- lm(borrower_rate ~ amount_funded + prosper_rating + listing_term + listing_monthly_payment
          + prosper_score + listing_category_id + stated_monthly_income * income_verifiable
          + employment_status_description + occupation + months_employed + lender_indicator 
          + monthly_debt + current_delinquencies + delinquencies_last7_years + public_records_last10_years 
          + public_records_last12_months + credit_lines_last7_years + inquiries_last6_months 
          + current_credit_lines + bankcard_utilization + total_inquiries + total_trade_items 
          + satisfactory_accounts + now_delinquent_derog + was_delinquent_derog 
          + is_homeowner, data = loans_lm)
summary(m1)
```

Looking at the summary, we can see which factors are significant and which ones aren't exlcuding the insignificant factors yeilds the following model, which avoids overfitting.
```{r}
# Remove all insignificant factors
m2 <- lm(borrower_rate ~ amount_funded + prosper_rating + listing_term + listing_monthly_payment
          + stated_monthly_income * income_verifiable + employment_status_description + occupation 
          + inquiries_last6_months + bankcard_utilization + total_inquiries + satisfactory_accounts 
          + is_homeowner, data = loans_lm)
summary(m2)
```

## Logistic Regression
Now we can create our model which seeks to find the real factors that create a risk of default. First, copy over the data we'll need to our new dataframe:
```{r}
loans_glm <- loans_lm
loans_glm$loan_status <- loans$loan_status
```

Now filter our the loans that haven't been completed, one way or the other - we only want loans we either know were paid, or loans we know weren't.
```{r}
# Filter out ongoing loans
loans_glm <- loans_glm[loans_glm$loan_status != 1, ]
```

With that done, we can create a new indicator variable that takes value 1 if the loan was paid in full, and value 0 if the loan was not.
```{r}
# Create indicator variable for if loan is paid in full
loans_glm$loan_paid <- as.factor(ifelse(loans_glm$loan_status == 4, 1, 0))
```

Once this is done, create a logistic regression that relates the factors from before with whether or not the loan is paid. 
```{r}
# Create logistic model for whether loan is paid in full
m3 <- glm(loan_paid ~ amount_funded + prosper_rating + listing_term + listing_monthly_payment
          + prosper_score + listing_category_id + stated_monthly_income * income_verifiable
          + employment_status_description + occupation + months_employed + lender_indicator 
          + monthly_debt + current_delinquencies + delinquencies_last7_years + public_records_last10_years 
          + public_records_last12_months + credit_lines_last7_years + inquiries_last6_months 
          + current_credit_lines + bankcard_utilization + total_inquiries + total_trade_items 
          + satisfactory_accounts + now_delinquent_derog + was_delinquent_derog 
          + is_homeowner, data = loans_glm, family = "binomial")
summary(m3)
```

As with before, remove the insignificant factors, yielding a model that identifies what aspects are actually important in deciding probability of default. 
```{r}
# Remove insignificant factors
m4 <- glm(loan_paid ~ amount_funded + prosper_rating + listing_term 
          + stated_monthly_income * income_verifiable + lender_indicator + monthly_debt 
          + public_records_last12_months, data = loans_glm, family = "binomial")
summary(m4)

```

## Analysis

The linear regression model suggests interest rates on loans depend on the following factors (that is, the following factors determine whether lenders consider you a credit risk): \
- amount_funded \
- prosper_rating \
- listing_term \
- listing_monthly_payment \
- stated_monthly_income * income_verifiable \
- employment_status_description \
- occupation \
- inquiries_last6_months \
- bankcard_utilization \
- total_inquiries \
- satisfactory_accounts \
- is_homeowner

Our analysis suggests loan defaults actually depend on a different set of factors, and that lenders' intuitions about what constitutes a risk are often imprecise or outright wrong. The real factors that play a role are as follows: \
- amount_funded \
- prosper_rating \
- listing_term \
- stated_monthly_income * income_verifiable \
- lender_indicator \
- monthly_debt \
- public_records_last12_months

This means the current way of deciding interest rates does not correspond directly to the likelihood of default, as one would expect. This is a market inefficiency, which could be exploited to produce a profit for a lender who worked with currently overcharged and underserved populations, such as non homeowners. 



